---
title: "Acoustic vs. Electronic Singer-Songwriters"
author: "Jona Benja Bosman"
output: 
  flexdashboard::flex_dashboard:
        storyboard: true
---

```{r setup, message=FALSE}
library(flexdashboard)
library(tidyverse)
library(plotly)
library(Cairo)
library(spotifyr)
library(ggplot2)
library(plyr)
library(dplyr)
source('spotify.R')
```

```{r}
#' Get a tidy audio analysis from Spotify.
#'
#' spotifyr returns Spotify's audio analysis as a large list. This function
#' uses list columns to create a structure that works more richly within the
#' tidyverse.
get_tidy_audio_analysis <- function(track_uri, ...) 
{
    get_track_audio_analysis(track_uri, ...) %>% 
        list %>% transpose %>% as_tibble %>% 
        mutate_at(vars(meta, track), . %>% map(as_tibble)) %>% 
        unnest(meta, track) %>% 
        select(
            analyzer_version,
            duration,
            contains('fade'),
            ends_with('confidence'),
            bars:segments) %>% 
        mutate_at(
            vars(bars, beats, tatums, sections), 
            . %>% map(bind_rows)) %>% 
        mutate(
            segments =
                map(
                    segments,
                    . %>% 
                        transpose %>% as_tibble %>% 
                        unnest(.preserve = c(pitches, timbre)) %>% 
                        mutate(
                            pitches = 
                                map(
                                    pitches, 
                                    . %>% 
                                        flatten_dbl %>% 
                                        set_names(
                                            c( 
                                                'C', 'C#|Db', 'D', 'D#|Eb', 
                                                'E', 'F', 'F#|Gb', 'G',
                                                'G#|Ab', 'A', 'A#|Bb', 'B'))),
                            timbre = 
                                map(
                                    timbre,
                                    . %>% 
                                        flatten_dbl %>% 
                                        set_names(
                                            c(
                                                'c1', 'c2', 'c3', 'c4', 
                                                'c5', 'c6', 'c7', 'c8',
                                                'c9', 'c10', 'c11', 'c12'))))))
}
```


```{r}
#' Normalise vectors for Computational Musicology.
#'
#' We use a number of normalisation strategies in Computational Musicology.
#' This function brings them together into one place, along with common
#' alternative names.
compmus_normalise <- compmus_normalize <- function(v, method = "euclidean")
{
    ## Supported functions
    
    harmonic  <- function(v) v * sum(1 / abs(v))
    manhattan <- function(v) v / sum(abs(v))
    euclidean <- function(v) v / sqrt(sum(v^2))
    chebyshev <- function(v) v / max(abs(v))
    clr       <- function(v) {lv <- log(v); lv - mean(lv)}
    
    ## Method aliases
    
    METHODS <-
        list(
            harmonic  = harmonic,
            manhattan = manhattan,
            L1        = manhattan,
            euclidean = euclidean,
            L2        = euclidean,
            chebyshev = chebyshev,
            maximum   = chebyshev,
            aitchison = clr,
            clr       = clr)
    
    ## Function selection
    

    if (!is.na(i <- pmatch(method, names(METHODS))))
        METHODS[[i]](v)
    else 
        stop('The method name is ambiguous or the method is unsupported.')
}

#' Compute pairwise distances for Computational Musicology in long format.
#'
#' We use a number of distance measures in Computational Musicology.
#' This function brings them together into one place, along with common
#' alternative names. It is designed for convenience, not speed.
compmus_long_distance <- function(xdat, ydat, feature, method = "euclidean")
{
    
    feature <- enquo(feature)
    
    ## Supported functions
    
    manhattan <- function(x, y) sum(abs(x - y))
    euclidean <- function(x, y) sqrt(sum((x - y) ^ 2))
    chebyshev <- function(x, y) max(abs(x - y))
    pearson   <- function(x, y) 1 - cor(x, y)
    cosine    <- function(x, y)
    {
        1 - sum(compmus_normalise(x, "euc") * compmus_normalise(y, "euc"))
    }
    angular   <- function(x, y) 2 * acos(1 - cosine(x, y)) / pi
    aitchison <- function(x, y)
    {
        euclidean(compmus_normalise(x, "clr"), compmus_normalise(y, "clr"))
    }
    
    ## Method aliases
    
    METHODS <-
        list(
            manhattan   = manhattan,
            cityblock   = manhattan,
            taxicab     = manhattan,
            L1          = manhattan,
            totvar      = manhattan,
            euclidean   = euclidean,
            L2          = euclidean,
            chebyshev   = chebyshev,
            maximum     = chebyshev,
            pearson     = pearson,
            correlation = pearson,
            cosine      = cosine,
            angular     = angular,
            aitchison   = aitchison)
    
    ## Function selection
    
    if (!is.na(i <- pmatch(method, names(METHODS))))
        bind_cols(
            crossing(
                xdat %>% select(xstart = start, xduration = duration),
                ydat %>% select(ystart = start, yduration = duration)),
            xdat %>% select(x = !!feature) %>% 
                crossing(ydat %>% select(y = !!feature)) %>% 
                transmute(d = map2_dbl(x, y, METHODS[[i]])))
    else 
        stop('The method name is ambiguous or the method is unsupported.')
}
```


```{r}
#' Gathers chroma vectors into long format.
#'
#' Gathers chroma vectors into long format for Computational Musicology.
compmus_gather_chroma <- function(data)
{
    data %>% 
    mutate(pitches = map(pitches, bind_rows)) %>% unnest(pitches) %>% 
    gather("pitch_class", "value", C:B) %>% 
    mutate(pitch_class = fct_shift(factor(pitch_class), 3))
}
```


Acoustic vs. Electronic Singer-Songwriters
-----------------------------------------------------------------------
### Introduction {data-commentary-width=500}

**Introduction**

This assignment will be about the differences and similarities between electronic and acoustic singer-songwriters. Acoustic and electronic music are a world apart, but it seems now that they can be closer than expected by the fact that both acoustic and electronic singer-songwriters exist. I define acoustic singer-songwriters as 1 or max. 2 people that sing and play exclusively on 1 acoustic instrument, mainly guitar or piano. Examples are Passenger and Ed Sheeran. Electronic singer-songwriters, to me, are people who produce their own music containing (not necessarily but also not limited to) synthesizers, looping stations or drums computers, and are able to perform it live. The music must contain roughly the same amount of singing as the acoustic songs. Examples are Chet Faker and JAIN. 

The question that will be examined in this report is: **Apart from the instruments used, what are the major differences between music of acoustic and electronic singer-songwriters?**

```{r echo = F}
acoustic <- get_playlist_audio_features('11122498650', '0a2aX4LlpTkeYLuTrVUu88')
electronic <- get_playlist_audio_features('11122498650', '2cANvNRpfzoZ6rkoaJQg25')

alles <- rbind(acoustic, electronic)

num_electronic <- nrow(electronic)
num_acoustic <- nrow(acoustic)
```

***

**Corpera**

There are two corpera that will be compared with each other. These are Spotify playlists. For the acoustic singer-songwriters, a playlist generated by Spotify that was called: **"Acoustic Singer-songwriters"** was chosen. The playlist was checked manually and all the songs satisfied the criteria stated above. This corpus represents the acoustic singer songwriters pretty well. There are **`r num_acoustic`** songs in this playlist.

<br/>

The playlist for the **electronic singer-songwriters** was made by a Spotify user and complemented by me.  It contains many songs that fulfill the criteria from above, but also more songs that belong to the Indie genre, with lots of electric guitars, but no electronic instruments. This playlist does not totally represent the playlist needed to execute this research, but I am planning on adjusting it manually. There are **`r num_electronic`** songs in this playlist.

### Mode

```{r echo = F}
mode_plot <- ggplot(alles, aes(x = playlist_name, fill = mode)) +
  geom_bar() +
  labs(title = "Mode", x = "Playlist", y = "No. of songs") 
```

```{r echo = F}
ggplotly(mode_plot)
```

***

The biggest difference between the two playlists is the mode of the songs. In the acoustic playlist, `r num_acoustic / 33 * 100 ` percent of the songs is in minor key. 
In the electronic playlist, `r num_electronic / 83 * 100 ` percent of the songs is in minor. That is a large difference. That difference is displayed in the barplots shown in this graph. 

### Energy

```{r echo = F}
energy_plot <- ggplot(alles, aes(x = playlist_name, y = energy)) +
  geom_boxplot() + 
  labs(title = "Energy", x = "Playlist", y = "Energy") 

ggplotly(energy_plot)

```

***

### Danceability

```{r}
danceability_plot <- ggplot(alles, aes(x = playlist_name, y = danceability)) +
  geom_boxplot() + 
  labs(title = "Danceability", x = "Playlist", y = "Danceability") 

ggplotly(danceability_plot)

```

***

The difference in danceability is smaller than one would expect. It seems logical that acoustic song are less danceable than electronic songs, mainly because of the electronic drums in electronic songs. But in the acoustic corpus, there are a few songs that are very acoustic, but performed by bands. That could influence the mean danceability value of the whole playlist.


***


### Chromogram 'Euphoria' by Dawson Hollow

```{r}
euphoria_chroma <- 
    get_tidy_audio_analysis('2R8K3yqNyi0B2JyuCdiaje') %>% 
    select(segments) %>% unnest(segments) %>% 
    select(start, duration, pitches)
```

```{r}
euphoria_chroma %>% 
    mutate(pitches = map(pitches, compmus_normalise, 'chebyshev')) %>% 
    compmus_gather_chroma %>% 
    ggplot(
        aes(
            x = start + duration / 2, 
            width = duration, 
            y = pitch_class, 
            fill = value)) + 
    geom_tile() +
    labs(x = 'Time (s)', y = NULL, fill = 'Magnitude') +
    theme_minimal()
```

***

This song is from the acoustic playlist and it indeed sounds very acoustic, but the energy is higher than one would expect and therefore, it is an outlier on the energy graph. This is probably due to the Irish-sounding violin that is in the song. That does give it a lot of energy despite it being an acoustic song.

This chromagram shows the chroma features of the song.

### Conclusion

**Conclusion**

The playlists are more alike than different, but there are some interesting features to explore, mainly the differences in mode and energy. The difference in danceability is 

*** 

**Suggestions for improvement **

  


