---
title: "Acoustic vs. Electronic Singer-Songwriters"
author: "Jona Benja Bosman"
output: 
  flexdashboard::flex_dashboard:
        storyboard: true
---

```{r setup, message=FALSE}
library(flexdashboard)
library(tidyverse)
library(plotly)
library(Cairo)
library(spotifyr)
library(ggplot2)
library(plyr)
library(dplyr)
source('spotify.R')
```

Acoustic vs. Electronic Singer-Songwriters
-----------------------------------------------------------------------
### Introduction {data-commentary-width=500}

**Introduction**

This assignment will be about the differences and similarities between electronic and acoustic singer-songwriters. Acoustic and electronic music are a world apart, but it seems now that they can be closer than expected by the fact that both acoustic and electronic singer-songwriters exist. I define acoustic singer-songwriters as 1 or max. 2 people that sing and play exclusively on 1 acoustic instrument, mainly guitar or piano. Examples are Passenger and Ed Sheeran. Electronic singer-songwriters, to me, are people who produce their own music containing (not necessarily but also not limited to) synthesizers, looping stations or drums computers, and are able to perform it live. The music must contain roughly the same amount of singing as the acoustic songs. Examples are Chet Faker and JAIN. 

The question that will be examined in this report is: **Apart from the instruments used, what are the major differences between music of acoustic and electronic singer-songwriters?**

```{r echo = F}
acoustic <- get_playlist_audio_features('11122498650', '0a2aX4LlpTkeYLuTrVUu88')
electronic <- get_playlist_audio_features('11122498650', '2cANvNRpfzoZ6rkoaJQg25')

alles <- rbind(acoustic, electronic)

num_electronic <- nrow(electronic)
num_acoustic <- nrow(acoustic)
```

***

**Corpera**

There are two corpera that will be compared with each other. These are Spotify playlists. For the acoustic singer-songwriters, a playlist generated by Spotify that was called: **"Acoustic Singer-songwriters"** was chosen. The playlist was checked manually and all the songs satisfied the criteria stated above. This corpus represents the acoustic singer songwriters pretty well. There are **`r num_acoustic`** songs in this playlist.

<br/>

The playlist for the **electronic singer-songwriters** was made by a Spotify user and complemented by me.  It contains many songs that fulfill the criteria from above, but also more songs that belong to the Indie genre, with lots of electric guitars, but no electronic instruments. This playlist does not totally represent the playlist needed to execute this research, but I am planning on adjusting it manually. There are **`r num_electronic`** songs in this playlist.

### Graphics

```{r echo = F}
mode_plot <- ggplot(alles, aes(x = playlist_name, fill = mode)) +
  geom_bar() +
  labs(title = "Mode", x = "Playlist", y = "No. of songs") 
```

```{r echo = F}
ggplotly(mode_plot)
```

***

Comments on plot

### Other plot

```{r echo = F}
danceability_plot <- ggplot(alles, aes(x = playlist_name, fill = mode)) +
  geom_bar() +
  labs(title = "Mode", x = "Playlist", y = "No. of songs") 
```

```{r echo = F}
ggplotly(danceability_plot)
```


***


### Chromogram Everything'll be Alright (Will's Lullaby)' by Joshua Radin

***

One really typical song in the acoustic playlist is **'Everything'll be Alright (Will's Lullaby)'** by *Joshua Radin*. The most remarkable value is danceability (0.53). The song is very soft and sweet and doesn't sound very danceable to me. A value of 0.53 seems higher than I would expect. I think 0.5 is a nice middle value for danceability. A lot of songs have this value. The value for acousticness is 0.693. Since this is a very acoustic sounding song, I would've expected the song to have higher acoustic value, but compared to the other songs, 0.693 is already pretty high. The mean acousticness of the acoustic playlist is 0.646 (SD = 0.268), so it is higher than average.



### Chromogram 'Dynabeat' by JAIN

***

In the electronic playlist, a typical song is **'Dynabeat'** by *JAIN*. The song is a lot more danceable than 'Everything'll be Alright (Will's Lullaby)' and that shows in the analysis (0.911). The acousticness is very low (0.0787). The instrumentalness is 0, which would imply that there are no instrumental parts in the song, but there is a instrumental solo in the bridge, so this analysis is not totally correct. 


### Conclusion
The playlists are more alike than different, but there are some interesting features to explore, mainly the differences in mode and energy. The difference in 
